# Chapter 6: Dimensionality reduction techniques

### Data wrangling and performing an analysis of longitudinal data

#### *Work of week 49 (02.12. - 08.12.2019)*

***

The data wrangling of two data sets (bprs & rats) was performed before. The script of the data wrangling exercise can be found [here](https://github.com/rikulampinen/IODS-project/blob/master/data/meet_and_repeat.R).  


## 1. Load and check the bprs_l and rats_l data sets


## 1.1. Load the data sets (bprs_l & rats_l)

```{r, packages, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load necessary packages
library(tidyr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(GGally)
library(knitr)
library(kableExtra)
library(stringr)
library(ggfortify)
```

```{r, loaddata, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load the wide data sets
bprs <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep = " ", header = TRUE)

rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE)

# load the long data sets
bprs_l <- read.table("C:/Users/richla/OneDrive/1 C - R-Folder/11-IODS-course/IODS-project/data/bprs_l.txt")
rats_l <- read.table("C:/Users/richla/OneDrive/1 C - R-Folder/11-IODS-course/IODS-project/data/rats_l.txt")

# check the data sets
str(bprs_l) 
str(rats_l)

# bprs data --> factor treatment & subject
bprs_l$treatment <- factor(bprs_l$treatment)
bprs_l$subject <- factor(bprs_l$subject)

# rats data --> factor ID & Group
rats_l$ID <- factor(rats_l$ID)
rats_l$Group <- factor(rats_l$Group)

# check the data again
glimpse(bprs_l)
glimpse(rats_l)
```
  
The categorical variables were not set as factors anymore, so I factorized them again. See the upper code.
  
***  
  
## 1.2.The data sets (bprs_l & rats_l)

```{r, checkdata, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# bprs_l data set
knitr::kable(bprs_l, caption = "BPRS - long data set") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "300px")

# rats_ data set
knitr::kable(rats_l, caption = "RATS - long data set") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "300px")
```
  
***
  
## 2. Analysis of the data sets (bprs_l & rats_l)

The data set analyses are swapped. So we run the data analyses of MABS book chapter 8 on the "rats_l" data set and the analyses of MABS book chapter 9 on the "bprs_l" data set.

## 2.1 Analyses of Chapter 8 of MABS using the RATS data (rats_l)

### 2.1.1. A first plot

```{r, rats_lanalysis1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# rats_l data set

# plot 1
ggplot(rats_l, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:16, times = 11)) + # needs to be checked!!!
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(rats_l$Weight), max(rats_l$Weight))) + 
  ggtitle("RATS - long data")
```

? What to do with + scale_linetype_manual(values = rep(1:10, times=4))

### 2.1.2. Standardized data

```{r, rats_lanalysis2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
head(rats_l)

# Standardise the variable Weight
rats_l <- rats_l %>%
  group_by(Time) %>%
  mutate(StdWeight = ((Weight - mean(Weight))/sd(Weight))) %>%
  ungroup()

# Glimpse the data
glimpse(rats_l)

# Plot again with the standardised bprs
ggplot(rats_l, aes(x = Time, y = StdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:16, times=11)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "Standardized Weight") +
  ggtitle("Standardized Weight")

```
  
### 2.1.3. Summary graphs

```{r, rats_lanalysis3, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
str(rats_l)

# Number of Time, baseline (Time = 1) included
n_rats <- rats_l$Time %>% unique() %>% length()
n_rats
# Summary data with mean and standard error of rats_l by group and time 
rats_s <- rats_l %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = (sd(Weight)/sqrt(n_rats))) %>%
  ungroup()

# Glimpse the data
str(rats_s)

# Plot the mean profiles
ggplot(rats_s, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line(col = "darkgreen") +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=2) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.8) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

### 2.1.4. Find the oulier

```{r, rats_lanalysis4, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
str(rats_l)

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 1).
rats_l10s <- rats_l %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(rats_l10s)

# Draw a boxplot of the mean versus group
ggplot(rats_l10s, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "darkgreen") +
  scale_y_continuous(name = "mean(Weight), Time 8-64")

# Create a new data by filtering the outliers and adjust the ggplot code the draw the plot again with the new data
rats_l10s1 <- rats_l10s %>%
    filter(mean < 550) %>%
      ungroup()
str(rats_l10s1)

# Draw a boxplot of the mean versus group with the outlier filtered data
ggplot(rats_l10s1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "darkgreen") +
  scale_y_continuous(name = "mean(Weight), Time 8-64")
```

### 2.1.4. T-test and ANOVA


```{r, rats_lanalysis4, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
str(rats_l10s1)

# Two-sample t-test
t.test(mean ~ Group == c(1,3), data = rats_l10s1) # that needs to be checked !!!!!!!!!!!!!!!!!!!

# Add the baseline from the original data as a new variable to the summary data
rats_l10s2 <- rats_l10s %>%
  mutate(baseline = rats$WD1)

rats_l10s2

# Fit the linear model with the mean as the response 
rats_fit <- lm(mean ~ baseline + Group, data = rats_l10s2)

# Compute the analysis of variance table for the fitted model with anova()
anova(rats_fit)
```
Check the results on the 










