# Chapter 5: Dimensionality reduction techniques

### Data wrangling and performing principal component analysis (PCA) & multiple correspondence analysis (MCA) 

#### *Work of week 48 (25.11. - 01.12.2019)*

***

## 1. Analysis of human data set

The data wrangling was done in two steps:  a data frame combination of two data sets and as a second step a refining the data to those variables we want to analyse. The data wrangling scrip was uploaded to my GitHub repository. You can find the data wrangling script [here](https://github.com/rikulampinen/IODS-project/blob/master/data/create_human.R).   
  
## 1.1. Load the data set

```{r, packages, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load necessary packages
library(tidyr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(GGally)
library(knitr)
library(kableExtra)
library(stringr)
library(ggfortify)
library(factoextra)
```

```{r, loaddata, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load the data set "human"
human <- read.table(file = 
             "C:/Users/richla/OneDrive/1 C - R-Folder/11-IODS-course/IODS-project/data/human_analys.txt")
```
  
*** 
  
## 1.2. The data set (overview and correlations)

The original complete data set from the United Nations was wrangled and combined into the so called "human" data set.
This human data set consists of 8 variables with 155 observations. The data set includes following variables:

* *country*         --> Name of the country 
* *ratio.sec.edu*   --> Ratio of second education of female/male  
* *ratio.lab.force* --> Ratio of labour forced female/male  
* *edu.expect*      --> Expected years of schooling  
* *life.exp*        --> Life expectancy at birth 
* *GNI*             --> Gross National Income per capita  
* *mat.mor.r*       --> Maternal mortality ratio  
* *adol.birth*      --> Adolescent birth rate  
* *rep.parliament*  --> Percetange of female representatives in parliament  

The data was collected from the United Nations. More information about the data and how it was collected can be found [here](http://hdr.undp.org/en/content/human-development-index-hdi).
Technical notes about calculating the human development indices can be found [here](http://hdr.undp.org/sites/default/files/hdr2015_technical_notes.pdf).  
   
Here you see the structure, the complete table and the summary of the wrangled human data set.

```{r, checkdata1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# check the data set "human"
str(human)

#  data set table
knitr::kable(human) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(width = "100%", height = "300px")

# data summary
knitr::kable(summary(human)) %>% 
  kable_styling(bootstrap_options = "striped", position = "center", font_size = 11)
```
  
***
  
### 1.2.1. Graphical overview of the human data set  

```{r, checkdata2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# visualization of human data set
ov_human <- ggpairs(human, mapping = aes(), title ="Overview of the human data set", 
                     lower = list(combo = wrap("facethist", bins = 20)), 
                     upper = list(continuous = wrap("cor", size = 3)))
ov_human
```
  
The overview plot shows the data distributions of all variables in the data set and its correlations to each other. Most of the data is basically normally distributed. GNI, maternal mortality rate  and adolescence birth rate show left skewed distribution representing a increased cound on low rates. On the upper part the correlations of the variables are shown as number values. Following you see the correlation matrix and a correlation plot presenting the values graphically.  

```{r, datacorr, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# calculate the correlation matrix and round it
cor_human <- cor(human) %>% round(digits = 2)

cor_human %>% knitr::kable(caption = "Correlation table of human data set") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# Specialized the insignificant value according to the significant level
p.mat <- cor.mtest(cor_human)$p

# visualize the correlation matrix
# correlations / colour shows the correlation values
corrplot(cor_human, method="pie", type="lower",  tl.cex = 0.65, p.mat = p.mat, sig.level = 0.01, tl.srt = 45, title="Correlations of the human data set", mar=c(0,0,1,0))  
```
  
This correlation plot shows nicely which variable corrate positively or negatively to each other.
The crossed square show a correlation which is not significant (p-value > 0.01). Strong positive correlations are seen between live expectation and expected years of education, gross national income with expected years of education and live expectation and adolescent birth rate with maternal mortality. So a better education leads to a higher income and so a higher life expectation. Maternal mortality rate depends on the rate of adolescence birth rate.  
  
***
  
## 1.3. Data Analysis

### 1.3.1. Principal component analysis (PCA) on the **not** standardized human data set 
  
Generally: in a PCA the data is transformed into new features called the principal components. The first principal component (PC) captures the maximum amount of variance of the features in the original data. The second PC captures the maximum amount of variability left and is orthogonal to the first PC (in a right angle to the first PC). All PCAs are uncorrelated analysis methods. Here we are using PC1 and PC2 for the data interpretation. 

Here the principal component analysis of the not standardized human data set, a table with the values of the principle components 1 & 2 and the PCA biplot.
  
```{r, not_std_PCA1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# perform PCA on not standardized data set
pca_human <- prcomp(human)

# PCA results on PC1 and PC2 on all variables
round(pca_human$rotation[,1:2], digits = 4) %>% knitr::kable(caption = "PCA result on all variables") %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# summary of the pca
s1 <- summary(pca_human)

# round the percentages of variance captured by the pc
pca_prc1 <- round(100 * s1$importance[2, ], digits = 3) 

# Principal components 1 & 2 table
round(100 * s1$importance[2, 1:2], digits = 5) %>% knitr::kable(caption = "Principle components") %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# prepare a label for the biplot
pca_label1 <-  paste0("GNI (", pca_prc1[1],"%)" )
pca_label2 <-  paste0("Maternal mortality (", pca_prc1[2],"%)" )

# draw the pca biplot
biplot(pca_human, choices = 1:2, cex = c(0.55, 0,9), col = c("grey40", "darkgreen"), xlab = pca_label1, 
                      ylab = pca_label2, main = "PCA biplot of non standardized human data set", margin =c(0,0,5,0)) 

```
  
**Let's try another kind of biplot**
```{r, not_std_PCA2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# more advanced biplot
autoplot(pca_human, data = human, label= TRUE, label.size = 3.0, colour = "darkgreen", loadings = TRUE, loadings.label = TRUE, loadings.colour = "red",) + ggtitle("PCA biplot of not standardized human data set") + xlab(paste0(pca_label1))  + ylab(paste0(pca_label2)) + theme_bw()
```
  
Principal component 1 represents 99.99% of the variance of the data set. PC2 captures just 0.01% of the variance. Pc1 is associated with the GNI variable. So, the Gross National Income per capita explains 99.99% of the principal component 1 of the data. 0.01 % is associated with maternal mortalility rate. This is the not standardized data set, that's why we get this result.

***
  
### 1.3.2. Principal component analysis (PCA) on the **standardized** human data set 
  
First we are scaling the data set so that the human is standardized so that the mean of all values is 0 and the standard diviation reaches 1. See here the summary table of the standardized data set.  

```{r, stdzd_human, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# scaling of human data set
human_stzd <- scale(human)

# summary of standardized human data set
knitr::kable(summary(human_stzd)) %>% 
  kable_styling(bootstrap_options = "striped", position = "center", font_size = 11)
```
  
In the summary table you can see that all mean values are 0. So the scaling of the human data set was successful.  
Now the PCA of this standardized data follows.

```{r, stdzd_human_PCA1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# perform PCA on standardized data set
pca_human_stzd <- prcomp(human_stzd)

# Principal components 1 & 2 table
round(pca_human_stzd$rotation[, 1:2], digits = 4) %>% knitr::kable(caption = "PC1 & 2 of standardized human data") %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# PC labels
s2 <- summary(pca_human_stzd)
pca_prc2 <- round(100 * s2$importance[2, ], digits = 1) 

pca_label2.1 <- paste0("Education and health (",pca_prc2[1],"%) ")
pca_label2.2 <- paste0("Female social participation (",pca_prc2[2],"%) ")

# Principal components 1 & 2 table
round(100 * s2$importance[2, 1:2], digits = 2) %>% knitr::kable(caption = "Principle components") %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# draw the PCA biplot
biplot(pca_human_stzd, choices = 1:2, cex = c(0.5, 0,9), col = c("grey40", "deeppink2"), xlab = pca_label2.1, ylab = pca_label2.2, main = "PCA biplot of standardized human data set", margin =c(0,0,5,0))
```


**Also here another kind of biplot**
```{r, stdzd_human_PCA2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# more advanced biplot
autoplot(prcomp(human_stzd), data = human, colour = "darkgreen", label= TRUE, label.size = 3.0, loadings = TRUE, loadings.label = TRUE, loadings.colour = "red",) + ggtitle("PCA biplot of standardized human data set") + xlab(paste0(pca_label2.1))  + ylab(paste0(pca_label2.2)) + theme_bw()
```
  
The principal component representing 16.2% of the data variation is strongly associated with parliament representatives and the labour force ratio (female / male). So the occurance of of woman in the labour market and in the political representation is mostly associated with principal component 2. 
The variables which show the strongest association with principal component 1 (representing 53.6% of the data variation) are life expectation, the school period, the ratio of female/male of second education, adolenscence birthrate and maternal mortality. So overall principal component 1 represents the health and education situation of the countries.
  
The PCA results differ a lot! To perform the PCA we need to make sure that the obsvervations are standardized, which means that the values are transformed in compareable values.  
We can see that in the PCA of the non-standardized human data set the principal component 1 represents almost all varation of the data set - so the Gross National Income per capita is the value representing principle component 1.  
The PCA of standardized values looks quite different. You can see that the PC1 & PC2 values spread all over the biplot and more variables are associated to the two principal components.  
  
The angle between the arrows represents the original features of the data set and we can interpret it as the correlation between these features. Is the angle between the arrows small, then there is a positive correlation.  
Also the angle between the feature and the principal component can be interpreted as the correlation between these two components. Again, if the angle is small there is a positive correlation. The length of the arrows is proportional to the standard deviation of the features.   
  
***  

## 2. Multiple Correspondence Analysis (MCA) 

## 2.1. Tea consumption behaviour MCA  

The "tea" data set includes data about tea consumption. Following, you see the MCA of the tea data set as it was performed on the DataCamp exercise.

### 2.1.1. Load the data set & check it  

```{r, factominer, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# load the package and the tea data set
library(FactoMineR)
data(tea)

# check the tea data set
str(tea)
```

***

### 2.1.2. Data overview

```{r, tea1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# column names to keep in the dataset
keep_tea <- c("Tea", "How", "how", "sugar", "where", "lunch") # from DataCamp

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_tea))

gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar(fill = "darkgreen") + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```
  
***
  
### 2.1.3. MCA

```{r, tea1mca1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# multiple correspondence analysis
mca_tea_time <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca_tea_time)

# visualize MCA
plot(mca_tea_time, invisible=c("ind"), habillage = "quali")
```
  
***

### 2.1.4. MCA, different biplot using "factoextra" package  

Here the MCA presented in another MCA biplot, performed with the function 'fviz_mca_biplot()' which is available from the 'factoextra' package. Here I added as  a individual value colour the cosine-squared (cos2), a value between 0 & 1. As closer to 1 this value gets as better it is projected on the dimension.
```{r, tea1mca2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
#draw a MCA biplot using factoextra function "fviz_mca_biplot"

fviz_mca_biplot(mca_tea_time, col.ind = "cos2", col.var = "red", label = "var", geom =  c("point","text"), labelsize = 4, arrows = c(FALSE,TRUE)) + labs(title = "MCA of tea consumption behaviour") + theme_grey() + theme(axis.line = element_line(size = 0.5), panel.background = element_rect(fill = "gray93"))
```
  
***

## 2.2. Another MCA analysis of different "tea" variables (reasons for drinking tea)

I want to know which reasons for drinking tea people have. So I have choosen following six variables to analyse by MCA:  

* *sex* --> Gender of the tea consumer
* *age_Q* --> Age class of the tea consumer
* *spirituality* --> Spiritual reason to drink tea
* *healthy* --> Health reason to drink tea
* *slimming* --> Tea consumer wants to loose weight or not
* *relaxing* --> Relaxation effect on tea on consumer or not
  
I prepare a separate tea_reason data set and perform the MCA of that data aliquote.

### 2.2.1. Prepare the data set and overview graph

```{r, tea2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}
# column names to keep for my analysis
keep_reason <- c("sex", "age_Q", "spirituality", "healthy", "slimming", "relaxing")

# select the 'keep_columns' to create a new dataset
tea_reason <- select(tea, one_of(keep_reason))

gather(tea_reason) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar(fill = "darkviolet", alpha = 0.6) + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

### 2.2.2. MCA of the "tea_reason" data

```{r, tea2mca, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align="center"}

# multiple correspondence analysis
mca_tea_reason <- MCA(tea_reason, graph = FALSE)

# summary of the model
summary(mca_tea_reason)

# draw the MCA biplot
fviz_mca_biplot(mca_tea_reason, col.ind = "cos2", col.var = "red", alpha.var = "contrib", label = "var", geom =  c("point","text"), labelsize = 3.5, arrows = c(FALSE,TRUE)) + labs(title = "MCA of tea consumption reasons") + theme_grey() + theme(axis.line = element_line(size = 0.5), panel.background = element_rect(fill = "gray93"))
```
  
The biplot shows interesting results. Variable categories which are very close show higher similarity. The plot includes two colour schemes representing the "cos2" of the individuals (as closer to 1 as better it is projected on the dimension) and the contribution of the variables to the dimenstions("contrib"). The biplot gives following information:  

* Men in the age of 25-34 drink tea not for health or weight loss reasons
* People in the age class of 45-59 show similarities with no relaxation and no spirituality reason for drinking tea
* People in the age class of 15-24, spirituality and relaxation reason are close and thereby similar
* Women & age class of 60+ have similarities with drinking tea for health and weight loosing reasons  

***
