# Chapter 6: Analysis of longitudinal data

### Data wrangling and performing an analysis of longitudinal data

#### *Work of week 49 (02.12. - 08.12.2019)*

***

The data wrangling of two data sets (bprs & rats) was performed before. The script of the data wrangling exercise can be found [here](https://github.com/rikulampinen/IODS-project/blob/master/data/meet_and_repeat.R).  


## 1. Load and check the data sets

## 1.1. Load the wide (bprs & rats) and long (bprs_l & rats_l) data sets

```{r, packageload, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load necessary packages
library(tidyr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(ggthemes)
library(GGally)
library(knitr)
library(kableExtra)
library(stringr)
library(lme4)
```

```{r, loaddata2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# load the wide data sets
bprs <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep = " ", header = TRUE)

rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE)

# load the long data sets
bprs_l <- read.table("C:/Users/richla/OneDrive/1 C - R-Folder/11-IODS-course/IODS-project/data/bprs_l.txt")
rats_l <- read.table("C:/Users/richla/OneDrive/1 C - R-Folder/11-IODS-course/IODS-project/data/rats_l.txt")

# check the data sets
str(bprs_l) 
str(rats_l)
```
  
The data sets show that the categorical variables "treatmeant" and "subject"  in the BPRS data set and the categorical variables "Group" and "ID" in the RATS data set were read as numbers. So I factorized them again. See the code below.  

```{r, check1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# bprs data --> factor treatment & subject
bprs_l$treatment <- factor(bprs_l$treatment)
bprs_l$subject <- factor(bprs_l$subject)

# rats data --> factor ID & Group
rats_l$ID <- factor(rats_l$ID)
rats_l$Group <- factor(rats_l$Group)

# check the data again
str(bprs_l)
str(rats_l)
```
 
*** 
## 1.2.The long data sets (bprs_l & rats_l)  

**The BPRS data set:**    
The BPRS (brief psychiatric rating scale) was measured on 40 male subjects with 2 treatment groups. One BPRS measurement was done before treatment started (week0), then weekly measurements followed up to 8 weeks in total. The BPRS assesses according to Vehkalahti & Everitt, 2019 the level of 18 symptom constructs which are used to evaluate if patients have schizophrenia. The BRRS symptoms constructs are measured in rates from 1 (not present) to 7 (extremely severe).
The "long" BPRS data set `bprs_l` consists of 5 variables with 360 observations. These variables include:  
  
* *treatment* - the psychological treatment the male subjects (treatment 1 & 2)  
* *subject* - the male individuals identification number
* *weeks* - the factor variable showing the week of treatment and BPRS evaluation
* *bprs* - the bprs values
* *week* - the week number of the BPRS evaluation  
  
```{r, check2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# bprs_l data set
knitr::kable(bprs_l, caption = "BPRS - long data set") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "300px")
```  
  
    
**The RATS data set:**  
This data consists of measurements on how rats grow. There were 3 groups of rats defined which got a different diet and the body weight of each animal was measured over a period of 9 weeks. The research question is if the growth profiles of the different diet groups differ (Vehkalahti & Everitt ,2019). The "long" RATS data set `rats_l` consists of 5 variables with 176 observations. The variables include:  

* *ID* - the individual rat's identification number
* *Group* - the diet group (1, 2 & 3)
* *WD* - the factorial value of the time (the day when the rats' weights were measured)
* *Weight* - weight of each individual rat at the given day
* *Time* - the time point of the weight measurement (days)  
  
```{r, check3, echo=TRUE, results='markdown', message=FALSE, warning=FALSE}
# rats_ data set
knitr::kable(rats_l, caption = "RATS - long data set") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% 
  scroll_box(height = "300px")
```
  
***
  
## 2. Analysis of the long data sets (bprs_l & rats_l)

The data set analyses presented in the MABS book and the DataCamp platform are swapped. So we run the data analyses of MABS book chapter 8 on the "rats_l" data set and the analyses of MABS book chapter 9 on the "bprs_l" data set. 

## 3. Analyses of chapter 8 of MABS using the RATS data (rats_l)
 
The long `rats_l` data will be analysed according to the instructions from chapter 8 of MABS. We start as a first step with a graphical overview of the data.

### 3.1. A first plot

```{r, rats1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# plot 1
ggplot(rats_l, aes(x = Time, y = Weight,linetype = ID)) +
  geom_line(color = "darkgreen") +
  scale_linetype_manual(values = rep(1:6, times= 3)) + # there are 16 IDs (6 linetypes are available - so repeated 3 times)
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "right") +
  scale_y_continuous(limits = c(min(rats_l$Weight), max(rats_l$Weight))) + 
  ggtitle("RATS: long data set ")
```
  
  
Here the `rats_l` data is plotted. The x-axis with the "Time" represents the days of the weight measurements, the y-axis shows the weights of the individual rats at the different measurement times. The plot is splitted into the three different diet groups. What can be seen in the graph is that the rat's starting weights in group 2 & 3 were higher than in group 1 and that the weights in group 2 & 3 increased in quite high rates, while in group 1 it seems that the weights increased in a much smaller rate.  
The data also shows the *"tracking"* phenomena (acc. MABS book, Vehkalahti & Everitt, 2019) where rats with a higher starting weight tend to have a higher weight increase over time. To make this *tracking* seen more clearly we standardise the `rats_l` data set.

***

### 3.2. Standardized data

The `rats_l` data set is standardised by calculating the "StdWeight"with following formular:  $StdWeight = (Weight - Mean(Weight)) / SD (Weight)$  
The calculated StdWeight is added as a new variable to the `rats_l` data set. Then the StdWeight is plotted.

```{r, rats2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Standardise the variable Weight
rats_l <- rats_l %>%
  group_by(Time) %>%
  mutate(StdWeight = ((Weight - mean(Weight))/sd(Weight))) %>%
  ungroup()

# Glimpse the data
glimpse(rats_l)

# Plot again with the standardised rats_l
ggplot(rats_l, aes(x = Time, y = StdWeight, linetype = ID)) +
  geom_line(color = "darkred") +
  scale_linetype_manual(values = rep(1:6, times=3)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "StdWeight - standardized Weight values") +
  ggtitle("RATS: standardized Weight values") + 
  theme(panel.grid.major.y = element_line(colour = "grey40"))
```
  
  
The "StdWeight"-plot shows the standardized weight values over the measurement period. Here the "tracking" (higher starting weights tent to increase more compared to lower weights and stay on a higher value for the measurement period) can be seen much better.  

***
  
### 3.3. Summary graphs
  
First you see an overview of the rats weights over the whole measurement period with boxplots. The different diet groups are seen in different colors.
```{r, rats3, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# add the Time as a categorical variable "Time1"
rats_l <- rats_l %>%
  mutate(Time1 = factor(rats_l$Time)) %>%
  ungroup()

# check the data 
str(rats_l)

# prepare a boxplot of the "rats_l" data set
ggplot(data = rats_l, aes(x = rats_l$Time1, y = Weight, fill = rats_l$Group)) +
  geom_boxplot() +
  ylab("Weight") + 
  xlab("Weight measurement time points [days]") +
  ggtitle("Rat weights over the measurment period") +
  scale_fill_discrete(name = "Diet group") +
  theme(legend.position = "right") 
```
  
Now, the sample number `n_rats` of the individually measured rat weights are calculated (in total 11 weight measurements, so n = 11). Then a new table `rats_s` is prepared, where the calculated mean weight of all rats & the standard error of each diet group is added as a variable. The standard error is calculated via following formular: $SE = SD(Weight)/ \sqrt n$
This new data table is then presented in the next plot.

```{r, rats4, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Number of Time, baseline (Time = 1) included
n_rats <- rats_l$Time %>% unique() %>% length()

# Summary data with mean and standard error of rats_l by group and time 
rats_s <- rats_l %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = (sd(Weight)/sqrt(n_rats))) %>%
  ungroup()

# Glimpse the data
str(rats_s)

# Plot the mean profiles
ggplot(rats_s, aes(x = Time, y = mean, linetype = Group,  color = Group, shape = Group)) +
  geom_line(size = 0.6) +
  scale_linetype_manual(name = "Diet group", values = c(1,2,3)) +
  geom_point(size=1.5) +
  scale_shape_manual(name = "Diet group", values = c(16,17,18)) +
  scale_color_manual(name = "Diet group", values = c("red", "darkgreen", "blue")) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.8) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "Mean (Weight) +/- SE (Weight)") + 
  ggtitle("Mean weight profiles of the different diet groups")
```
  
This graph shows the rat weight development of the three different diet groups over the measurement period. Group 1 starts on a lower level and the weight increases just slightly (~250 to 280). Group 2 and group 3 weights start form a higher values and also the rats's weights decrease on a higher level (in group 2 an increase of more that 50 (+/- of course...), in group 3 around 50 (+/- a smaller value)). What you can see is that the standard errors in group 2 & 3 are much higher, so there is a much bigger weight difference between the individual rats within the diet group.  

***

### 3.4. Find the outlier

Here we will have now a look how the rat's weight developed after the first measurement. Therefore we prepare a new data table `rats_l10s` where the weight data without time = 1 is summarised by calculating the mean value of weight per diet group.

```{r, rats5, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 1).
rats_l10s <- rats_l %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(rats_l10s)
```
  
From this new data table we prepare a boxplot:
```{r, rats6, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Draw a boxplot of the mean versus group
ggplot(rats_l10s, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=3, fill = "black") +
  scale_y_continuous(name = "Mean (Weight) / Time 8-64 [days]") + 
  xlab("Diet group") + 
  scale_fill_discrete(name = "Diet group") +
  ggtitle("Mean weights of diet groups excluding the first measurement")
```
  
The boxplot shows the mean rat weights of each diet group of the measurements after the first measurement (day 1). The boxplots of all three diet groups actually show an outlier. These outliers can disturb conclusions from further comparisons of the diet groups. So we have to remove these outliers as I did it for each group seen in the code below. I filter out the outliers and create a new data table `rats_l10s1` where the mean values without the outliers are collected. Then a new boxplot of the diet groups is prepared.

```{r, rats7, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Create a new data by filtering the outliers and adjust the ggplot code the draw the plot again with the new data
rats_l10s1 <- rats_l10s %>%
    filter(
      (mean > 250 & Group == 1) | 
      (mean < 550 & Group == 2) |
      (mean > 500 & Group == 3))  %>% 
      ungroup()

str(rats_l10s1)

# Draw a boxplot of the mean versus group with the outlier filtered data
ggplot(rats_l10s1, aes(x = Group, y = mean, fill = Group)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "black") +
  scale_y_continuous(name = "Mean (Weight) / Time 8-64 [days]") +
  xlab("Diet group") + 
  scale_fill_discrete(name = "Diet group") +
  ggtitle("Mean weights of diet groups excluding the first measurement")
```
  
Here we see the boxplot graph with the weights means over the measurement period 8 to 64 days without outlier values. With that data we continue to analyse the diet group differences.

***

### 3.5. T-test and ANOVA

We cannot perform a t-test between the groups since we have 3 groups and not 2 groups.

```
t.test(mean ~ rats_l10s2$Group == 2, paired = TRUE) 
# Two-sample t-test group 1 & 3
t.test(mean ~ Group == 3, data = rats_l10s1) 
# cannot be done on three groups
```  
Now we continue with creating a new data table `rats_l10s2`where a so called "baseline" weight values are created from the first weigh measurement on day 1 (WD1). So each group weight baseline was defined. See the table below with the mean weight values of the whole measurement period and the starting weight

```{r, rats9, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Add the baseline from the original data as a new variable to the summary data
rats_l10s2 <- rats_l10s %>%
  mutate(baseline = rats$WD1)

# check the data table
rats_l10s2 %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center") %>% scroll_box(height = "300px")
```

Using this new data table with the mean weight values and the baseline weight values we perform an anova test to see which diet group had the highest weight increase over the measurement period. there are differences between the diet groups. In the following codes I run first a linear model comparing the group means to each other (from the data table `rats_l10s2` excluding the starting weight measurement) and in the second linear model I compare the baseline weights of each group with the mean weight value of each group (also from the `rats_l10s2` data table).  

I continue now to compare the diet group differences with performing an ANOVA and a post-hoc test (TukeyHSD) to compare the group means, since I want to know if the diet groups are different from each other. The ANOVA is performed from the data table `rats_l10s2`. This is the data table without the first measurement (start value) and inluding the outlier values.

```{r, rats10, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# linear regression model of mean values versus diet groups
rats_fit1 <- lm(mean ~ Group, data = rats_l10s2)
# regression model output
round(rats_fit1$coefficients, digits = 2) %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```
  
The model shows that the mean value of group 2 is 222g higher and the mean value of group 3 is 262g higher compared to group 1 mean weights. Let's have a look now if the diet groups differ to each other. With a post-hoc test after the ANOVA we can check the group means.

```{r, rats11, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Compute the analysis of variance table of rats_fit1 <- lm(mean ~ Group, data = rats_l10s2) with anova()
rats_anova1 <- anova(rats_fit1)
rats_anova1
```
  
The ANOVA of the linear model of mean weight ~ diet group shows that group 2 has a significant different weight increase compared to group 1 & group 3.

```{r, rats12, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# another kind of anova calculation --> compute the anova with aov()
rats_aov2 <- aov(mean ~ Group, data = rats_l10s2)
summary(rats_aov2)

# compare the anova results with tukeyHSD
TukeyHSD(rats_aov2, "Group", ordered = TRUE, conf.level = 0.95)
```
  
Here we can see that the secon ANOVA function gives the same result and this result can be put in the TukeyHSD function. This post-hoc test gives us then the differences between the diet groups. Group 1 & 2 are significantly different in the mean weight. Group 1 & 3 are as well. Group 2 and 3 are not significantly different in the weight means (there is the outlier influence).  
  
  
Now lets see what is the difference of each group to the baselines.  
  
```{r, rats13, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Fit the linear model with the mean as the response 
rats_fit2 <- lm(mean ~ baseline + Group, data = rats_l10s2)

# regression model output
round(rats_fit2$coefficients, digits = 2) %>% knitr::kable() %>% kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```
  
Here a regression model of the mean rats weights compared to the weight baseline (the first weight measurement of each dietary group). The results shows that the weight difference between baseline and group 2 is around 34g and group 3 and baseline is around 23g different. So group 2 has actually the highest weight increase rates over the measurement period.
  
```{r, rats14, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Compute the analysis of variance table of rats_fit2 <- lm(mean ~ baseline + Group, data = rats_l10s2) with anova()
rats_anova2 <- anova(rats_fit2)
rats_anova2
```
  
What is seen from the second ANOVA is that the baseline has a significant influence on the weight increase later in the different diet groups and that group 2 has a rather good significant weight increase relative to the baseline compared to group 1 and group 3. So it seems that diet 2 leads to the highest weight increase.
  
***
  
## 4. Analyses of Chapter 9 of MABS using the BPRS data (bprs_l)

### 4.1. A first plot

```{r, bprs_lanalysis1, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}

str(bprs_l)

#plotting the bprs_l data
ggplot(bprs_l, aes(x = week, y = bprs, group = treatment)) +
  geom_point() + 
  scale_x_continuous(name = "Weeks", breaks = seq(0,8)) + 
  scale_y_continuous(name = "BPRS value") + 
  theme(legend.position = "top") + 
  ggtitle("BPRS value overview")

#plotting the bprs_l data
ggplot(bprs_l, aes(x = week, y = bprs, linetype = subject)) +
  geom_line(color = "darkgreen") +
  scale_linetype_manual(values = rep(1:6, times = 4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Weeks", breaks = seq(0,8,1)) + 
  scale_y_continuous(name = "BPRS values observed", breaks = seq(10,100,5)) + 
  theme(legend.position = "bottom") + 
  ggtitle("BPRS: data values overview") +
  theme(legend.box.background = element_rect(),legend.box.margin = margin(2, 2, 2, 2))
```
  
***

### 4.2. Linear model of bprs data

```{r, bprs_lanalysis2, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}

# create a regression model bprs_l_reg
bprs_l_reg <- lm(bprs ~ week + treatment, data = bprs_l)

# print out a summary of the model
summary(bprs_l_reg)
```
  
  
*** 

### 4.3. Random Intercept Model

```{r, bprs_lanalysis3, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# Create a random intercept model
bprs_l_rim <- lmer(bprs ~ week + treatment + (1 | subject), data = bprs_l, REML = FALSE)

# Print the summary of the model
summary(bprs_l_rim)




# confidence interval --> 1.96 * standard error
# t value --> test if the value (+/- confidence interval can be 0)
# the mixed model (lmer) takes into account the correlations between the observations (here between the subjects) --> so it is more accurate --> the standard
# p-value can be calculated  by inverse function of t value --> a function to invert the t value to the p value!!
```
  

*** 

### 4.4. Random Intercept and Random Slope Model

```{r, bprs_lanalysis4, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}

# create a random intercept and random slope model
bprs_l_rim1 <- lmer(bprs ~ week + treatment + (week | subject), data = bprs_l, REML = FALSE)


# here the other times (weeks) are included - your get a regression model defined for each subject for each time period (for each week) - so here 9 (0-8 weeks) different regression models 
# so it inclueds


# print a summary of the model
summary(bprs_l_rim1)

# perform an ANOVA test on the two models
anova(bprs_l_rim1, bprs_l_rim)
```
  
***

### 4.5. Random Intercept and Random Slope Model with interaction

```{r, bprs_lanalysis5, echo=TRUE, results='markdown', message=FALSE, warning=FALSE, fig.align = 'center'}
# create a random intercept and random slope model
bprs_l_rim2 <- lmer(bprs ~ week * treatment + (week | subject), data = bprs_l, REML = FALSE)

# print a summary of the model
summary(bprs_l_rim2)

# perform an ANOVA test on the two models
anova(bprs_l_rim2, bprs_l_rim1)

# draw the plot of bprs_l
ggplot(bprs_l, aes(x = week, y = bprs, linetype = subject)) +
  geom_line(color = "darkgreen") +
  scale_linetype_manual(values = rep(1:6, times = 4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Weeks", breaks = seq(0,8,1)) + 
  scale_y_continuous(name = "BPRS values observed", breaks = seq(10,100,5)) + 
  theme(legend.position = "bottom") + 
  ggtitle("BPRS value overview")

# Create a vector of the fitted values
bprs_l_fit <- fitted(bprs_l_rim2)

# Create a new column "fitted_bprs" to bprs_l
bprs_l$fitted_bprs <- bprs_l_fit

# draw the plot of fitted bprs values
ggplot(bprs_l, aes(x = week, y = fitted_bprs, linetype = subject)) +
  geom_line(color = "darkred") +
  scale_linetype_manual(values = rep(1:6, times = 4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Weeks", breaks = seq(0,8,1)) + 
  scale_y_continuous(name = "BPRS values modeled", breaks = seq(10,100,5)) + 
  theme(legend.position = "bottom") + 
  ggtitle("BPRS value overview")
```



